---
title: "phase1_report"
author: "Noah Giebink"
date: "8/12/2019"
output:
  pdf_document: default
  html_document: default
---
#Objectives
1. Acquire, clean, and explore iNat and Adventure Scientists (AS) data
2. Identify candidate species and localities for initial SDM comparisons


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#packages
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(knitr)

#data
adventure_scientists_data_raw <-
  read.csv("data/iNaturalist_AdventureScientists.csv")
adventure_scientists_data_raw_expanded <-
  read.csv("data/iNaturalist_AdventureScientists_taxonInfoExpanded.csv")
iNat_montana <- 
  read.csv("data/iNat_montana.csv")
iNat_california <-
  read.csv("data/iNat_california.csv")
iNat_washington <-
  read.csv("data/iNat_washington.csv")
iNat_utah <-
  read.csv("data/iNat_utah.csv")
iNat_arizona <-
  read.csv("data/iNat_arizona.csv")
iNat_as <-
  read.csv("data/iNat_as.csv") 
#iNat_as is updated Adventure Scientists data as of 2019-8-6
#retrieved after iNat_[state] data (also 2019-8-6)
#to exclude AS data from iNat

#iNat states compiled
iNat_full <-
  rbind(iNat_arizona, iNat_california, iNat_montana, iNat_utah, iNat_washington) 
  
#remove Adventure Scientists observations from iNat
as_observations <- select(iNat_as, id)
iNat_only <- filter(iNat_full, !(iNat_full$id %in% as_observations$id))
```


###Top five species observed by state: AS
```{r, echo=FALSE}
#AS data
top_5 = adventure_scientists_data_raw_expanded %>%
  filter(place_state_name != "") %>%
  filter(place_state_name %in% c("Washington", "Utah", 
                                 "Montana", "California", 
                                 "Arizona")) %>%
  group_by(place_state_name, scientific_name) %>%
  summarize(num_records = n()) %>%
  ungroup() %>%
  group_by(place_state_name) %>%
  top_n(n = 5, wt = num_records) %>%
  arrange((place_state_name), desc(num_records)) 

kable(top_5)

```

###Top five species observed by state: iNat
```{r}
top_5_iNat <- iNat_only %>%
  group_by(place_state_name, scientific_name) %>%
  summarize(num_records = n()) %>%
  ungroup() %>%
  group_by(place_state_name) %>%
  top_n(n = 5, wt = num_records) %>%
  arrange((place_state_name), desc(num_records))

kable(top_5_iNat)
  
```

###Best candidate species
####Top five coverage in both AS and iNat
```{r}
top_shared <- top_5 %>%
  ungroup() %>%
  mutate(place_state_name = as.character(place_state_name), 
         scientific_name = as.character(scientific_name)) %>%
  inner_join(top_5_iNat %>%
               ungroup() %>%
               mutate(place_state_name = as.character(place_state_name),
                      scientific_name = as.character(scientific_name)), by = "scientific_name") %>%
  select(scientific_name, iNat_data_state = place_state_name.y, num_records_iNat_data = num_records.y, 
         as_data_state = place_state_name.x, num_records_as_data = num_records.x)

kable(top_shared)
```


##2. Identify candidate species and localities for initial SDM comparisons
Within the contiguous U.S., Adventure Scientists have been deployed in Arizona, California, Montana, Utah, and Washington. 


###Top five species observed in each state for both sampling methods
```{r, echo=FALSE}
speciesHeatmap <- ggplot(top_5,
                         aes(x = scientific_name,
                             y = place_state_name)) +
  geom_tile(aes(fill = num_records)) +
  coord_flip() +
  geom_text(aes(label = num_records)) +
  scale_fill_gradient(low = "white",
                      high = "red", name = "Number of records") +
  xlab("Scientific name") +
  ylab("State") +
  theme_bw() +
  theme(axis.text.y = element_text(face = "italic"),
        axis.text.x = element_text(angle = 45, 
                                   hjust = 1)) +
  ggtitle("Adventure Scientists")

speciesHeatmap_iNat_only <- ggplot(top_5_iNat,
                         aes(x = scientific_name,
                             y = place_state_name)) +
  geom_tile(aes(fill = num_records)) +
  coord_flip() +
  geom_text(aes(label = num_records)) +
  scale_fill_gradient(low = "white",
                      high = "red") +
  xlab("Scientific name") +
  ylab("State") +
  theme_bw() +
  theme(axis.text.y = element_text(face = "italic"),
        axis.text.x = element_text(angle = 45, 
                                   hjust = 1)) +
  ggtitle("iNat only")

both = ggarrange(speciesHeatmap, speciesHeatmap_iNat_only, common.legend = TRUE)
both
```

Sampling that targets remote areas yields remarkably different species observations than non-targeted, largely urban sampling. AS sampling yields more species diversity, evenness, or both. For AS, only one species was in the top five for more than one state; the rest are uniquely common to one state. iNaturalist sampling alone yields more repeated common species across states.
